---
title: "Cleaning Divvy Tripdata for a Whole Year"
author: "Omar Osman"
date: "2024-02-01"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Set the working directory
knitr::opts_knit$set(root.dir = 'C:/Users/Omar/Desktop/cyclistic_case_study')
```
# Introduction\

This document outlines the process of cleaning Divvy trip data for a whole year.\

## Cleaning Strategy\

1. **Drop Corrupted Rows**\
    a- Drop all rows with start time after end time.\
    b- Drop rows with missing start or end station data.\

2. **Convert `member_casual` Column**\
    - Convert `member_casual` column into a boolean table called `Member` containing `TRUE` for member and `FALSE` for casual.\

3. **Change `rideable_type` Column**\
    - Change `rideable_type` column to `is_electric`, converting its type from text to boolean.\

4. **Process Stations Data**\
    a- Create station tables for each dataset and populate them with valid station data only.\
    b- Merge stations with similar coordinates and names.\
    c- Fill empty station names using station name and coordinates.\
    d- Merge all station tables into one.\
    e- Minimize Merged Table rows.\

5. **Finalizing Data Processing**\
    a- Drop `station_id`.\
    b- Add new IDs columns to the main table.\
    c- Drop stations columns from the original table.\

6. **Drop Trips ID Column**\
    - For analysis quickness, drop trips ID column since they are all unique.\

7. **Create a new table and add each stations trip count to a new column alongside station id**\

8. **Export our final tables as csv**\
  
##Cleaning:\
#### include usesd packeges\
```{r include usesd packeges}
suppressMessages(library(DBI))
suppressMessages(library(RSQLite))
```

## Data Preparation.\

#### Create a list of all the used CSV files
```{r Intialise csv files paths}
csv_paths <- c(
  "data/orginal_data/202212-divvy-tripdata.csv",
  "data/orginal_data/202301-divvy-tripdata.csv",
  "data/orginal_data/202302-divvy-tripdata.csv",
  "data/orginal_data/202303-divvy-tripdata.csv",
  "data/orginal_data/202304-divvy-tripdata.csv",
  "data/orginal_data/202305-divvy-tripdata.csv",
  "data/orginal_data/202306-divvy-tripdata.csv",
  "data/orginal_data/202307-divvy-tripdata.csv",
  "data/orginal_data/202308-divvy-tripdata.csv",
  "data/orginal_data/202309-divvy-tripdata.csv",
  "data/orginal_data/202310-divvy-tripdata.csv",
  "data/orginal_data/202311-divvy-tripdata.csv"
)
```
#### Create a sqlite data base with all the tables:
```{r Creating SQLite database}
add_csv_to_sqlite <- function(csv_paths, db_path) {
  # Create a connection to the SQLite database
  con <- dbConnect(RSQLite::SQLite(), dbname = db_path)
  
  for (path in csv_paths) {
    # Extract the file name without the path
    file_name <- basename(path)
    
    # Apply rules to generate a suitable table name
    table_name <- gsub(".csv", "", file_name)
    if (grepl("^[0-9]", table_name)) {
      table_name <- paste0("T", table_name)
    }
    table_name <- gsub("-", "_", table_name)
    # Read the CSV file into a data frame
    df <- read.csv(path, na.strings = "")
    
    # Replace empty strings with NULL
    df[df == ""] <- NA
    
    # Write the data frame to a new table in the database
    dbWriteTable(con, table_name, df, overwrite = TRUE)
    
    # Print a message indicating that the table has been created
    cat(paste0("Table '", table_name, "' has been created in the database.\n"))
  }
  
  # Close the database connection
  dbDisconnect(con)
}
``` 
*Run the add_csv_to_sqlite function:*
```{r run add_csv_to_sqlite function}
# Specify the database path
db_path <- "data/all_data.db"

# Call the function with the CSV paths and the database path
add_csv_to_sqlite(csv_paths, db_path)

```

## 1-Drop corrupted rows:\

### 1-a-Drop all the rows that have Start time after End time:
```{r Drop all the rows that have Start time after End time}
drop_rows_and_report_db <- function(db_path) {
  # Create a connection to the SQLite database
  con <- dbConnect(RSQLite::SQLite(), dbname = db_path)
  
  # Get a list of table names in the database
  table_names <- dbListTables(con)
  
  for (table_name in table_names) {
    # Count the initial number of rows
    initial_rows <- dbGetQuery(con, paste0('SELECT COUNT(*) FROM "', table_name, '";'))$`COUNT(*)`
    
    # Construct an SQL command to delete rows where started_at is after ended_at
    sql_command <- paste0('DELETE FROM "', table_name, '" WHERE started_at > ended_at;')
    
    # Execute the SQL command
    dbExecute(con, sql_command)
    
    # Count the final number of rows
    final_rows <- dbGetQuery(con, paste0('SELECT COUNT(*) FROM "', table_name, '";'))$`COUNT(*)`
    
    # Calculate the number of rows dropped
    rows_dropped <- initial_rows - final_rows
    
    # Print a message indicating the number of rows dropped for each table
    cat(paste0("For table '", table_name, "', ", rows_dropped, " rows dropped due to started_at > ended_at.\n"))
  }
  
  # Close the database connection
  dbDisconnect(con)
}
```

*Run the drop_rows_and_report_db function:*

```{r Run the drop_rows_and_report_db function}
# Specify the existing database path
existing_db_path <- "data/all_data.db"

# Call the function with the existing database path
drop_rows_and_report_db(existing_db_path)
```

### 2-b-Drop rows that have no data about start or end station:

```{r Drop rows that have no data about start or end station}
drop_rows_if_null <- function(db_path) {
  # Create a connection to the SQLite database
  con <- dbConnect(RSQLite::SQLite(), dbname = db_path)
  
  # Get a list of table names in the database
  table_names <- dbListTables(con)
  
  for (table_name in table_names) {
    # Check if the table contains the required columns
    if (all(c("end_station_name", "end_station_id", "end_lat", "end_lng", "start_station_name", "start_station_id", "start_lat", "start_lng") %in% dbListFields(con, table_name))) {
      # Construct an SQL command to delete rows where any of the required columns are null
      sql_command <- paste0('DELETE FROM "', table_name, '" WHERE (end_station_name IS NULL AND end_station_id IS NULL AND end_lat IS NULL AND end_lng IS NULL) OR (start_station_name IS NULL AND start_station_id IS NULL AND start_lat IS NULL AND start_lng IS NULL);')
      
      # Count the initial number of rows
      initial_rows <- dbGetQuery(con, paste0('SELECT COUNT(*) FROM "', table_name, '";'))$`COUNT(*)`
      
      # Execute the SQL command
      dbExecute(con, sql_command)
      
      # Count the final number of rows
      final_rows <- dbGetQuery(con, paste0('SELECT COUNT(*) FROM "', table_name, '";'))$`COUNT(*)`
      
      # Calculate the number of rows dropped
      rows_dropped <- initial_rows - final_rows
      
      # Print a message indicating the number of rows dropped for each table
      cat(paste0("For table '", table_name, "', ", rows_dropped, " rows dropped due to null values.\n"))
    } else {
      # Print a message indicating that the table does not contain the required columns
      cat(paste0("Table '", table_name, "' does not contain all the required columns.\n"))
    }
  }
  
  # Close the database connection
  dbDisconnect(con)
}
```

*runing drop_rows_if_null function.*

```{r runing drop_rows_if_null function}
# Specify the existing database path
existing_db_path <- "data/all_data.db"

# Call the function with the existing database path
drop_rows_if_null(existing_db_path)
```

## 2-Convert `member_casual` Column.\
#### Convert member_casual column into a boolean table called Member and contains TRUE instead of member and False instead of casual.
```{r Convert member_casual column into a boolean table}
convert_members_col_to_bool <- function(db_path) {
  # Create a connection to the SQLite database
  con <- dbConnect(RSQLite::SQLite(), dbname = db_path)
  
  # Get a list of table names in the database
  table_names <- dbListTables(con)
  i=0
  for (table_name in table_names) {
    # Check if the table contains the required columns
    if (all(c("ride_id", "rideable_type", "member_casual","end_station_name", "end_station_id", "end_lat", "end_lng", "start_station_name", "start_station_id", "start_lat", "start_lng") %in% dbListFields(con, table_name))) {
      
      # Add a new boolean column 'is_member' based on 'member_casual'
      dbExecute(con, paste0("
        ALTER TABLE ", table_name, "
        ADD COLUMN is_member TEXT;"))
        
      dbExecute(con, paste0("
        UPDATE ", table_name, "
        SET is_member = (CASE WHEN member_casual = 'member' THEN 'TRUE'
                              WHEN member_casual = 'casual' THEN 'FALSE'
                              ELSE NULL END);
      "))
        
        
      
      # Drop the old column 'member_casual'
      dbExecute(con, paste0("
        ALTER TABLE ", table_name, "
        DROP COLUMN member_casual;
      "))
      
      #cat("Processed table:", table_name, "\n")
      # Check if i is divisible by 3 to perform VACUUM operation
      if (i %% 3 == 0) {
        #cat("Vacuuming the database...\n")
        dbExecute(con, "VACUUM")
        #cat("Database vacuumed.\n")
      }
      i <- i + 1
    }
  }
  cat("member_casual Column converted to Boolean")
  # Disconnect from the database
  dbDisconnect(con)
}
```

*running convert_members_col_to_bool function.*

```{r running convert_members_col_to_bool function.}
convert_members_col_to_bool("data/all_data.db")
```


## 3-Change `rideable_type` Column.\
#### Change rideable_type column to is_electric and change the type from Text to Boolean\
**electric_bike to True, docked_bike and classic_bike to False**\
I then decided to contact a Divvy data center representative. He politely followed back,\
explaining to me that "there are two kinds of bikes: electric and non-electric.\
The non-electric bikes are nowadays labelled as classic_bike and docked_bike.\
Hence, if you see docked_bike and classic_bike in the same dataset, \
know that you can categorize them together, as they are physically and functionally the same"\
```{r Change `rideable_type` Column.}
convert_rideable_col_to_bool <- function(db_path) {
  # Create a connection to the SQLite database
  con <- dbConnect(RSQLite::SQLite(), dbname = db_path)
  
  # Get a list of table names in the database
  table_names <- dbListTables(con)
  for (table_name in table_names) {
    # Check if the table contains the required columns
    if (all(c("ride_id", "rideable_type", "is_member","end_station_name", "end_station_id", "end_lat", "end_lng", "start_station_name", "start_station_id", "start_lat", "start_lng") %in% dbListFields(con, table_name))) {
      
      # Add a new boolean column 'is_electric' based on 'rideable_type'
      dbExecute(con, paste0("
        ALTER TABLE ", table_name, "
        ADD COLUMN is_electric TEXT;"))
        
      dbExecute(con, paste0("
        UPDATE ", table_name, "
        SET is_electric = (CASE WHEN rideable_type = 'electric_bike' THEN 'TRUE'
                                WHEN rideable_type =  'docked_bike'  THEN 'FALSE'
                                WHEN rideable_type =  'classic_bike'  THEN 'FALSE'
                              ELSE NULL END);
      "))
        
        
      
      # Drop the old column 'rideable_type'
      dbExecute(con, paste0("
        ALTER TABLE ", table_name, "
        DROP COLUMN rideable_type;
      "))
      
      #cat("Processed table:", table_name, "\n")
      # Check if i is divisible by 3 to perform VACUUM operation
    }
  }
  
  # Disconnect from the database
  dbDisconnect(con)
}
```

*runing convert_rideable_col_to_bool function.*

```{r runing convert_rideable_col_to_bool function.}
convert_rideable_col_to_bool("data/all_data.db")
```


## 4-Process Stations Data.\
(Clean then Split it from the main tables while keeping only the ID of the stations):\

### 4-a-Create station table for each table and populate it with valid station data only:\

create_station_tables function :\
  create a table named table_name+station for each data table:   \
    this table will have 4 rows (station_id,station_id,station_lng,station_lat)\
  The function will have 2 queries : 1 to select start station data and the other for the end station.\
  Finally it will drop all non unique rows in stations tables.
```{r create_station_tables function.}
create_station_tables <- function(db_file_path) {
  # Load SQLite database
  con <- dbConnect(RSQLite::SQLite(), dbname = db_file_path)
  
  # Get list of table names
  table_names <- dbListTables(con)
  for (table_name in table_names) {
    
    if (all(c("end_station_name", "end_station_id", "end_lat", "end_lng", "start_station_name", "start_station_id", "start_lat", "start_lng") %in% dbListFields(con, table_name))) {
    # Define new table name
    new_table_name <- paste0("`", table_name, "_station`")

    #Create a new table with 4 columns
    create_table_query <- paste0(
      "CREATE TABLE IF NOT EXISTS ", new_table_name, " AS ",
      "SELECT start_station_id AS station_id, start_station_name AS station_name, start_lng AS station_lng, start_lat AS station_lat FROM [", table_name, "]",
      "UNION ",
      "SELECT end_station_id AS station_id, end_station_name AS station_name, end_lng AS station_lng, end_lat AS station_lat FROM [", table_name, "]"      )
    dbExecute(con, create_table_query)
    
    count_query_before <- dbGetQuery(con, paste0('SELECT COUNT(*) FROM "', table_name, '";'))$`COUNT(*)`
    
    # Drop non-unique rows
    drop_duplicates_query <- paste0(
      "DELETE FROM ", new_table_name, " WHERE ROWID NOT IN ",
      "(SELECT MIN(ROWID) FROM ", new_table_name, " GROUP BY station_id, station_name, station_lng, station_lat)"
    )
    dbExecute(con, drop_duplicates_query)

    count_query_after <- dbGetQuery(con, paste0('SELECT COUNT(*) FROM "', table_name, '";'))$`COUNT(*)`
    cat(table_name,"had :",count_query_before, " and now have:",count_query_after,"\n")

    }
  }
  
  # Disconnect from the database
  dbDisconnect(con)
}


```

*running create_station_tables function.*

```{r running create_station_tables function.}
create_station_tables("data/all_data.db")
```

### 4-b-Merge stations with similar coordinates and names.\
Merge Same Stations Data.\

*After creating the stations table we have found that:*\
some coordinates are so close with same station name and id ,\
so we will merge them\
EX: 13276	Stockton Dr & Wrightwood Ave	-87.638737559	41.9313308\
    13276	Stockton Dr & Wrightwood Ave	-87.638737321	41.93136704\
    
So we will create minimize_station_table function to process newly created stations table,\
First it will create a minimized station table for each station table,\
Second Remove rows with station_id only, station_name only, or both when there is a corresponding row with both\
Then Remove duplicate rows based on station_id and station_name \
Finally it will drop old station tables.
```{r minimize_station_table function.}
minimize_station_table <- function(db_file_path) {
  # Load SQLite database
  con <- dbConnect(RSQLite::SQLite(), dbname = db_file_path)
  
  # Get list of table names
  table_names <- dbListTables(con)
  
  for (table_name in table_names) {
    if (all(c("station_id", "station_name", "station_lng", "station_lat") %in% dbListFields(con, table_name))) {
      # Define new table name
      new_table_name <- paste0("`", table_name, "_minimized`")
      
      # Create a new table with the same structure
      create_table_query <- paste0(
        "CREATE TABLE IF NOT EXISTS ", new_table_name, " AS ",
        "SELECT * FROM [", table_name, "]"
      )
      dbExecute(con, create_table_query)

      # Remove rows with station_id only, station_name only, or both when there is a corresponding row with both
      remove_partial_rows_query <- paste0(
        "DELETE FROM ", new_table_name, " WHERE (station_id IS NOT NULL AND (station_name IS NULL) OR ",
        "(station_name IS NOT NULL AND (station_id IS NULL))) AND EXISTS (",
        "SELECT 1 FROM ", new_table_name, " t2 WHERE t2.station_id = ", new_table_name, ".station_id AND t2.station_name = ", new_table_name, ".station_name ",
        "AND t2.station_id IS NOT NULL AND t2.station_name IS NOT NULL AND t2.station_lat IS NOT NULL AND t2.station_lng IS NOT NULL)"
      )
      dbExecute(con, remove_partial_rows_query)
      # Remove duplicate rows based on station_id and station_name and taking the median coordinates of each group
      remove_duplicates_query <- paste0(
                                          "DELETE FROM ", new_table_name, " WHERE ROWID NOT IN (",
                                          "  SELECT ROWID FROM (",
                                          "    SELECT ROWID, station_id, station_name,",
                                          "           AVG(station_lng) OVER (PARTITION BY station_id, station_name) AS median_lng,",
                                          "           AVG(station_lat) OVER (PARTITION BY station_id, station_name) AS median_lat",
                                          "    FROM ", new_table_name,
                                          "  )",
                                          "  WHERE ROWID IN (",
                                          "    SELECT MIN(ROWID) FROM ", new_table_name, " GROUP BY station_id, station_name",
                                          "  ))"
                                        )


      dbExecute(con, remove_duplicates_query)
      # Drop old table if it exists
      drop_table_query <- paste0(
        "DROP TABLE  ", table_name
      )
      dbExecute(con, drop_table_query)
    

    }
  }
  
  # Disconnect from the database
  dbDisconnect(con)
}



```

*runing minimize_station_table function.*

```{r runing minimize_station_table function.}

db_file_path <- "data/all_data.db"
minimize_station_table(db_file_path)

```


### 4-c-Fill empty station names using station name and coordinates.\
Fill empty stations names for starting and ending stations using station name and coordinates .\
fill_station_names_exact_value -> is a function that will fill the name using id or exact coordinates.\
```{r fill_station_names_exact_value function.}
fill_station_names_exact_value <- function(databaseName, tableName, id_col_name, station_col_name, lat_col_name, lng_col_name) {
  # Connect to the database
  db <- dbConnect(SQLite(), databaseName)

query <- paste0(
  "UPDATE ", tableName,
  " SET ", station_col_name, " = (",
  " SELECT station_name ",
  " FROM ",tableName,"_station_minimized ",
  " WHERE (", tableName, ".", id_col_name, " = ",tableName,"_station_minimized.station_id) ",
  "  OR (", tableName, ".", lat_col_name, " = ",tableName,"_station_minimized.station_lat AND ", tableName, ".", lng_col_name, " = ",tableName,"_station_minimized.station_lng)",
  ")",
  " WHERE ", station_col_name, " IS NULL"
)

# Execute the query
dbExecute(db, query)
  
  # Disconnect from the database
  dbDisconnect(db)
}
fill_station_names_approx_value <- function(databaseName, tableName, id_col_name, station_col_name, lat_col_name, lng_col_name,approx) {
  # Connect to the database
  db <- dbConnect(SQLite(), databaseName)
  
  # Construct the SQL query
  query <- paste0(
    "UPDATE ", tableName,
    " SET ", station_col_name, " = (",
    " SELECT station_name ",
    " FROM ",tableName,"_station_minimized",
    " WHERE (ABS(",tableName, ".", lat_col_name, " - ",tableName,"_station_minimized.station_lat )+ABS( ", tableName, ".", lng_col_name, " -",tableName,"_station_minimized.station_lng) <",approx,
    ")) ",
    "WHERE ", station_col_name, " IS NULL"
  )

  # Execute the query
  dbExecute(db, query)
  # Disconnect from the database
  dbDisconnect(db)
}
```
fill_station_names_approx_value -> will fill the name using approximate coordinates.\
```{r complete_station_table_names function}

complete_station_table_names <- function(database_location) {
  con <- dbConnect(RSQLite::SQLite(), dbname = database_location)
  
  i <- 1
  # Helper function to avoid repetition
  fill_station_approx_values <- function(threshold) {
    fill_station_names_approx_value(database_location, table_name, "start_station_id", "start_station_name", "start_lat", "start_lng", threshold)
    fill_station_names_approx_value(database_location, table_name, "end_station_id", "end_station_name", "end_lat", "end_lng", threshold)
  }
  
  # Get the names of all tables in the database
  tables <- dbListTables(con)
  
  # Iterate over all tables
  for (table_name in tables) {
    # Get the column names of the current table
    col_names <- dbListFields(con, table_name)

    # Check if the current table has the given columns for start station
    if ("start_station_name" %in% col_names & "start_station_id" %in% col_names & "start_lat" %in% col_names & "start_lng" %in% col_names) {
      #cat("Processing table:", table_name, "start station\n")
      fill_station_names_exact_value(database_location, table_name, "start_station_id", "start_station_name", "start_lat", "start_lng")

      # Call the helper function with different threshold values
      thresholds <- c(0.001, 0.005, 0.01, 0.05,0.1)
      for (threshold in thresholds) {
        fill_station_approx_values(threshold)
      }
    }
    
    # Check if the current table has the given columns for end station
    if ("end_station_name" %in% col_names & "end_station_id" %in% col_names & "end_lat" %in% col_names & "end_lng" %in% col_names) {
      #cat("Processing table:", table_name, "end station\n")
      fill_station_names_exact_value(database_location, table_name, "end_station_id", "end_station_name", "end_lat", "end_lng")

      # Call the helper function with different threshold values
      thresholds <- c(0.001, 0.005, 0.01, 0.05,0.1)
      for (threshold in thresholds) {
        fill_station_approx_values(threshold)
      }

      # Count number of rows after modifications
    }
     if (i %% 3 == 0) {
        #cat("Vacuuming the database...\n")
        dbExecute(con, "VACUUM")
        #cat("Database vacuumed.\n")
      }
      i <- i + 1
  }
  
  # Close the database connection
  dbDisconnect(con)
}
```

*running fill_station_names_exact_value function.*

```{r running fill_station_names_exact_value function.}
complete_station_table_names("data/all_data.db")
```
  
  


  

### 4-d-Merge all station tables into one.\  
Merge all the tripdata_station_minimized columns in 1 table.\
and drop all the tripdata_station_minimized tables. 
```{r merge_and_drop_tables function}
merge_and_drop_tables <- function(db_file_path) {
  # Load SQLite database
  con <- dbConnect(SQLite(), db_file_path)
  
  # Get list of table names
  table_names <- dbListTables(con)
  
  # Create a list to store the names of minimized tables
  minimized_tables <- c()
  
  for (table_name in table_names) {
    if (all(c("station_id", "station_name", "station_lng", "station_lat") %in% dbListFields(con, table_name))) {
      minimized_tables <- c(minimized_tables, table_name)
    }
  }
  
  # Merge all minimized tables into one
  merge_tables_query <- paste0(
    "CREATE TABLE IF NOT EXISTS \"merged_tripdata_station_minimized\" AS ",
    "SELECT * FROM ", paste(sprintf('"%s"', minimized_tables), collapse = " UNION ALL SELECT * FROM ")
  )
  dbExecute(con, merge_tables_query)
  
  # Drop all minimized tables
  for (minimized_table in minimized_tables) {
    drop_table_query <- paste0("DROP TABLE IF EXISTS ", sprintf('"%s"', minimized_table))
    dbExecute(con, drop_table_query)
  }


  # Disconnect from the database
  dbDisconnect(con)
}
```
  
*runing merge_and_drop_tables function.*

```{r runing merge_and_drop_tables function.}
merge_and_drop_tables("data/all_data.db")

```
  
  
  
### Minimize Merged Table rows
```{r}
minimize_merged_table <- function(db_file_path) {
  # Load SQLite database
  con <- dbConnect(SQLite(), db_file_path)

  #Drop duplicated rows in merged table
  drop_duplicates_query <- "DELETE FROM merged_tripdata_station_minimized WHERE ROWID NOT IN (SELECT MIN(ROWID) FROM merged_tripdata_station_minimized  GROUP BY station_id, station_name, ROUND(station_lat, 3), ROUND(station_lng, 3))"
  dbExecute(con, drop_duplicates_query)
  drop_duplicates_query <- paste0(
                                          "DELETE FROM  merged_tripdata_station_minimized WHERE ROWID NOT IN (",
                                          "  SELECT ROWID FROM (",
                                          "    SELECT ROWID, station_id, station_name,",
                                          "           AVG(station_lng) OVER (PARTITION BY station_id, station_name) AS median_lng,",
                                          "           AVG(station_lat) OVER (PARTITION BY station_id, station_name) AS median_lat",
                                          "    FROM merged_tripdata_station_minimized",
                                          "  )",
                                          "  WHERE ROWID IN (",
                                          "    SELECT MIN(ROWID) FROM merged_tripdata_station_minimized GROUP BY station_id,",
                                          "station_name  ))"
                                        )

  # Remove rows with station_id only, station_name only, or both when there is a corresponding row with both
  #delete rows from the merged_tripdata_station_minimized table that have incomplete data. Specifically, it deletes rows where either the station_id or station_name is not null but the other is null, or where either the station_lng or station_lat is null. Additionally, it deletes rows where there exists another row with the same station_id, station_name, and non-null station_lat and station_lng values 1.
  remove_partial_rows_query <- paste0(
    "DELETE FROM merged_tripdata_station_minimized
    WHERE 
    (station_id IS NOT NULL AND (station_name IS NULL) OR
     (station_name IS NOT NULL AND (station_id IS NULL)) OR
     (station_lng IS NULL OR (station_lat IS NULL)))
    AND EXISTS (
        SELECT 1 
        FROM merged_tripdata_station_minimized t2 
        WHERE 
          t2.station_name = merged_tripdata_station_minimized.station_name 
          AND t2.station_id IS NOT NULL 
          AND t2.station_name IS NOT NULL 
          AND t2.station_lat IS NOT NULL 
          AND t2.station_lng IS NOT NULL
    );
"
  )
  dbExecute(con, remove_partial_rows_query)

  # Remove rows with the same station_id, station_name, and small differences in lat/lng
   remove_similar_rows_query <- paste0(
                                      "CREATE TABLE stations AS
                                        SELECT 
                                            station_id,
                                            station_name,
                                            station_lng,
                                            station_lat
                                        FROM (
                                            SELECT 
                                                station_id,
                                                station_name,
                                                station_lng,
                                                station_lat,
                                                ROW_NUMBER() OVER (PARTITION BY station_id, station_name ORDER BY ROWID) as row_num
                                            FROM merged_tripdata_station_minimized
                                        ) AS ranked
                                        WHERE row_num = 1;"
                                        
                                          )

  dbExecute(con, remove_similar_rows_query)
  # Remove rows with the same station_id, station_name, and small differences in lat/lng
  drop_old_stations_table <- "DROP TABLE IF EXISTS merged_tripdata_station_minimized"
  dbExecute(con, drop_old_stations_table)

  #clean up null names
  clean_up_null_names_query <- ("DELETE FROM stations WHERE station_name IS NULL")
  dbExecute(con, clean_up_null_names_query)
  # Disconnect from the database
  dbDisconnect(con)
}


```

```{r}
minimize_merged_table("data/all_data.db")
```

## 5-Finalizing Data Processing.**\

### 5-a- Drop rows that have null stations name either in (start station or end station)
```{r}
drop_null_station_names_rows <- function(db_file_path) {
  # Load SQLite database
  con <- dbConnect(RSQLite::SQLite(), dbname = db_file_path)
  
  # Get list of table names
  table_names <- dbListTables(con)
  for (table_name in table_names) {
    
    if (all(c("end_station_name", "end_station_id", "end_lat", "end_lng", "start_station_name", "start_station_id", "start_lat", "start_lng") %in% dbListFields(con, table_name))) {
      sql_command <- paste0('DELETE FROM "', table_name, '" WHERE (end_station_name IS NULL OR start_station_name IS NULL);')

      # Execute the SQL command
      dbExecute(con, sql_command)
      
    }
  }
  dbDisconnect(con)
}
```
```{r}
drop_null_station_names_rows("data/all_data.db")
```

  
  
## Finalizing data processing after merging\

### Drop station_id 
```{r}
delete_stations_id_column <- function(db_path) {
  # Connect to the SQLite database
  con <- dbConnect(SQLite(), dbname = db_path)
  
  dbExecute(con, "ALTER TABLE stations DROP COLUMN station_id")

  # Disconnect from the database
  dbDisconnect(con)
}

```
```{r}
delete_stations_id_column("data/all_data.db")
```


### add the new ids columns to the all_trips_data table
```{r add_new_ids_columns function}
add_new_ids_columns <- function(db_path) {
  # Connect to the SQLite database
  con <- dbConnect(SQLite(), dbname = db_path)

  table_names <- dbListTables(con)
  i <- 1
  
  for (table_name in table_names) {
    
    if (all(c("end_station_name", "end_station_id", "end_lat", "end_lng", "start_station_name", "start_station_id", "start_lat", "start_lng") %in% dbListFields(con, table_name))) {
      # Step 1: Add new columns to the original table
      dbExecute(con, paste0("ALTER TABLE ", table_name, " ADD COLUMN start_station_key_id INTEGER"))
      dbExecute(con, paste0("ALTER TABLE ", table_name, " ADD COLUMN end_station_key_id INTEGER"))
      
      # Step 2: Update the new columns based on the mapping with the stations table
      dbExecute(con, paste0("
                 UPDATE ", table_name, "
                 SET start_station_key_id = (SELECT rowid FROM stations WHERE stations.station_name = ", table_name, ".start_station_name),
                     end_station_key_id = (SELECT rowid FROM stations WHERE stations.station_name = ", table_name, ".end_station_name)
                 "))
      
      #cat("Processing table: ", table_name, " finished\n")
      
      # Check if i is divisible by 3 to perform VACUUM operation
      if (i %% 3 == 0) {
        #cat("Vacuuming the database...\n")
        dbExecute(con, "VACUUM")
        #cat("Database vacuumed.\n")
      }
      i <- i + 1
    }
  }
  
  # Disconnect from the database
  dbDisconnect(con)
}


```
 
*runing add_new_ids_columns function.*

```{r runing add_new_ids_columns function.}
add_new_ids_columns("data/all_data.db")
```

### Drop stations columns from the original table after it was already refrenced using new id
```{r function drop_trips_columns.}
drop_trips_columns <- function(db_path){
  # Connect to the SQLite database
  con <- dbConnect(SQLite(), dbname = db_path)

  table_names <- dbListTables(con)
  i <- 1
  
  for (table_name in table_names) {
    
    if (all(c("end_station_name", "end_station_id", "end_lat", "end_lng", "start_station_name", "start_station_id", "start_lat", "start_lng") %in% dbListFields(con, table_name))) {
      # List of columns to drop
      columns_to_drop <- c("end_station_name", "end_station_id", "end_lat", "end_lng", 
                            "start_station_name", "start_station_id", "start_lat", "start_lng")
    
      # Loop through each column to drop
      for (column_name in columns_to_drop) {
        query <- paste0("ALTER TABLE ", table_name, "  DROP COLUMN ", column_name)
        dbExecute(con, query)}

      #cat("Processing table: ", table_name, " finished\n")
      
      # Check if i is divisible by 3 to perform VACUUM operation
      if (i %% 3 == 0) {
        #cat("Vacuuming the database...\n")
        dbExecute(con, "VACUUM")
        #cat("Database vacuumed.\n")
      }
      i <- i + 1
    }
  }

  # Disconnect from the database
  dbDisconnect(con)
}
```

*running function drop_trips_columns.*

```{r running function drop_trips_columns.}
drop_trips_columns("data/all_data.db")
```


## Merge all data tables\

```{r merge_data_tables function.}
merge_data_tables <- function(db_file_path) {
  new_table_name <- "all_trips_data"
  
  # Load SQLite database
  con <- dbConnect(RSQLite::SQLite(), dbname = db_file_path)

  # Get a list of all tables in the database
  all_tables <- dbListTables(con)

  # Initialize a vector to store the tables that meet the criteria
  valid_tables <- character(0)

  # Iterate over all tables
  for (table_name in all_tables) {
    # Check if the specified columns exist in the table
    required_columns <- c("ride_id", "is_electric")

    if (all(required_columns %in% dbListFields(con, table_name))) {
      # If columns exist, add the table to the valid_tables vector
      valid_tables <- c(valid_tables, table_name)
    }
  }

  # Check if any valid tables were found
  if (length(valid_tables) > 0) {
    # Create a new table and insert data from the valid tables
    select_queries <- lapply(valid_tables, function(table) {
      paste("SELECT * FROM", table)
    })
    
    query <- paste0("CREATE TABLE ", new_table_name, " AS ", 
                    paste(select_queries, collapse = " UNION ALL "))
    dbExecute(con, query)
    
    cat("Tables merged successfully. New table created:", new_table_name, "\n")
    # Drop old tables
    for (table in valid_tables) {
      dbExecute(con, paste0("DROP TABLE ", table))
      cat("Table dropped:", table, "\n")
    }
  } else {
    cat("No tables with the required columns found.\n")
  }
  cat("Vacuuming the database...\n")
  dbExecute(con, "VACUUM")
  cat("Database vacuumed.\n")
  # Close the database connection
  dbDisconnect(con)
}
```

*running merge_data_tables function.*

```{r running merge_data_tables function.}
merge_data_tables("data/all_data.db")
```
### Check that all the trips ID are unique.\

```{r Check_Id_uniqueness function.}
Check_Id_uniqueness <- function(db_file_path) {

  # Load SQLite database
  con <- dbConnect(RSQLite::SQLite(), dbname = db_file_path)
  query <- ("SELECT ride_id
            FROM all_trips_data
            GROUP BY ride_id
            HAVING COUNT(ride_id) > 1;")
  print(dbGetQuery(con, query))
  
  dbDisconnect(con)
}
```

*runing Check_Id_uniqueness function.*

```{r runing Check_Id_uniqueness function.}
Check_Id_uniqueness("data/all_data.db")
```

## 6- Drop Trips ID Column.\
**Since all the  trips IDs are unique and if we needed to use an id we can use rowid**\
```{r drop_ride_id_col Function}
drop_ride_id_col <- function(db_file_path) {

  # Load SQLite database
  con <- dbConnect(RSQLite::SQLite(), dbname = db_file_path)
  query <- ("ALTER TABLE all_trips_data DROP COLUMN ride_id")
  dbExecute(con, query)
  cat("Vacuuming the database...\n")
  dbExecute(con, "VACUUM")
  cat("Database vacuumed.\n")
  dbDisconnect(con)
}
```

*running drop_ride_id_col Function.* 

```{r running drop_ride_id_col Function}
drop_ride_id_col("data/all_data.db")
```
## 7-Create a new table and add each stations trip count to a new column alongside station id.\
*since its comlicated to count trips count in tableau.*\
```{r create_trips_count_table function.}
create_trips_count_table <- function(db_file_path) {
  con <- dbConnect(RSQLite::SQLite(), dbname = db_file_path)
  
  create_StationTotalTrips_table_query <- "CREATE TABLE IF NOT EXISTS StationTotalTrips (
    StationRowId INTEGER,
    TotalTrips INTEGER
  );"
  dbExecute(con, create_StationTotalTrips_table_query)
  
  fill_StationTotalTrips_table_query <- "INSERT INTO StationTotalTrips (StationRowId, TotalTrips)
    SELECT
      s.rowid AS StationRowId,
      COALESCE(dep.Departures, 0) + COALESCE(arr.Arrivals, 0) AS TotalTrips
    FROM
      stations s
    LEFT JOIN
      (SELECT
        start_station_key_id,
        COUNT(*) AS Departures
      FROM
        all_trips_data
      GROUP BY
        start_station_key_id) dep ON s.rowid = dep.start_station_key_id
    LEFT JOIN
      (SELECT
        end_station_key_id,
        COUNT(*) AS Arrivals
      FROM
        all_trips_data
      GROUP BY
        end_station_key_id) arr ON s.rowid = arr.end_station_key_id;"
  
  dbExecute(con, fill_StationTotalTrips_table_query)
  
  dbDisconnect(con)
  
  cat("StationTotalTrips table has been created and filled with data.\n")
}
```

*running create_trips_count_table function.*

```{r running create_trips_count_table function.}
create_trips_count_table("data/all_data.db")
```
## 8-Export our final tables as csv.\
*so we can use them in tableau public.*\
```{r export_tables_as_csv function.}
export_tables_as_csv <- function(db_file_path, output_dir) {
  con <- dbConnect(RSQLite::SQLite(), dbname = db_file_path)
  
  # Export StationTotalTrips table
  station_total_trips <- dbGetQuery(con, "SELECT * FROM StationTotalTrips")
  write.csv(station_total_trips, file.path(output_dir, "StationTotalTrips.csv"), row.names = FALSE)
  cat("StationTotalTrips table exported successfully.\n ")
  
  # Export all_trips_data table
  all_trips_data <- dbGetQuery(con, "SELECT * FROM all_trips_data")
  write.csv(all_trips_data, file.path(output_dir, "all_trips_data.csv"), row.names = FALSE)
  cat("all_trips_data table exported successfully.\n ")
  
  # Export stations table
  stations <- dbGetQuery(con, "SELECT rowid, * FROM stations")
  write.csv(stations, file.path(output_dir, "stations.csv"), row.names = FALSE)
  cat("stations table exported successfully.\n ")
  
  #Vacuuming the database.
  dbExecute(con, "VACUUM")
  cat("Database vacuumed.\n")
  
  dbDisconnect(con)
  
}

```

*running export_tables_as_csv function.*\

```{r running export_tables_as_csv function.}
export_tables_as_csv("data/all_data.db","data/final_csv_files/")
```

